Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness

### 2017-08-31 manuscript/FinalDraft_rev/ReviewResponses.tex

\documentclass[ucm,12pt]{ucletter}

\setlength\parindent{1cm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{xcolor}

%\name{Justin D. Yeakel}
\telephone{(209)~285-9571}
\email{jyeakel@ucmerced.edu}

\newcounter{section}
\newcounter{subsection}[section]
\setcounter{secnumdepth}{3}
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
                                   {-3.5ex \@plus -1ex \@minus -.2ex}%
                                   {2.3ex \@plus.2ex}%
                                   {\normalfont\Large\bfseries}}
\newcommand\subsection{\@startsection{subsection}{2}{\z@}%
                                     {-3.25ex\@plus -1ex \@minus -.2ex}%
                                     {1.5ex \@plus .2ex}%
                                     {\normalfont\large\bfseries}}
\renewcommand \thesection{\@arabic\c@section}
\renewcommand\thesubsection{\thesection.\@arabic\c@subsection}

\makeatother

\begin{document}
\begin{letter}{
    Philosophical Transactions of the Royal Society B\\
    Publishing Section\\
    The Royal Society\\
    6-9 Carlton House Terrace\\
    London SW1Y 5AG\\
    England\\
    \centerline{\bf{Re: Eco-evolutionary dynamics and collective migration:}}\\
     \centerline{\bf{implications for salmon metapopulation robustness}}
}

\opening{To the editorial committee,}

\setstretch{1.2}

\section*{Reviewer I}
\subsection*{Major comments:}
\textcolor{cyan}{Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect. In the second line of Eq (4), the immigrant term $(1-w_i) R_i\[\mu_j\]$ is independent of $\mu_i$, so that the derivative of this term is zero. This is as if selection did not change the allele frequencies/mean trait value of the immigrant subpopulation, which is obviously untrue (there is strong selection on the locally maladapted immigrants, favouring those who happen to be closer to the local optimum). One cannot apply Lande's equation without adapting it to the immigration model.}

We appreciate the Reviewer's concern, however note that equation 4) is written as
\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
\end{align}
and because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to the strength and direction of selection.

\textcolor{cyan}{
The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference. Ronce and Kirkpatrick focused on equilibria (such as the dominant vs subordinate state here ) and on local adaptation, not metapopulation resilience/stability, but did study the consequence of disturbances for switching between alternative equilibria (see the next point).
}

ff

\textcolor{cyan}{
The bifurcation the authors call a fold bifurcation is not a fold bifurcation. Figure 1 suggests a pitchfork bifurcation (which is degenerate but will appear in symmetric models). An alternative possibility is that there exists an asymmetric equilibrium with a "dominant state" and a "subordinate state" also at low m, next to the symmetric equilibrium shown at low m in figure 1, so that the system has two stable equilibria; here the "FB" point in figure 1 would be the point where the symmetric equilibrium is destabilized (this is the pattern found by Ronce and Kirkpatrick). This second alternative seems likely to me because the abrupt changes seen in figure 1 and also in figure 5 may well indicate an attractor switch. The nature of the bifurcation should be properly investigated (one can find Jacobians etc with numerical analysis for the deterministic system).
}

ff

\textcolor{cyan}{
The ms does not describe how recovery time was measured. The deterministic system returns to an equilibrium asymptotically, i.e., in infinite time. For practice, one defines a return ``close enough" for recovery. The details of this should be given, otherwise the results are not reproducible. Moreover, I think that near the bifurcation point total population size will equilibrate fast, but the distribution of individuals over the two habitats takes a longer time to equilibrate (the system is on the edge between a symmetric and an asymmetric equilibrium). If time to recovery includes the time that $N_1$ and $N_2$ individually reach their steady states, then this can explain the peak of recovery time at the bifurcation; but this does not mean that the total population size takes long to recover. If total population size was monitored, but recovery was interpreted as "return to original" whereas the system attains a different equilibrium (symmetric vs asymmetric, see above), then obviously recovery time will be infinite, but again this does not mean that the population does not recover from near extinction.
}

ff

\textcolor{cyan}{
A strong "portfolio effect" as measured by PE in equation 6 indicates resilience against metapopulation extinction if PE is high due to a low CV of total population size. But PE is also high if the local populations have high CV. As I argued above, near the bifurcation N_1 and N_2 may be much less stable than $N_T=N_1+N_2$, i.e., the peak of PE may have little to do with the resilience of total population size.
}

hehehe

\subsection*{Minor comments:}
\textcolor{cyan}{``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}

\textcolor{cyan}{The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes. Generally, the authors should pay attention to reproducibility.}

\section*{Reviewer II}
\subsection*{Major comments}
My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported. What is presented as a Fold bifurcation is in reality a Pitchfork bifurcation. Pitchfork bifurcations are dependent on perfect symmetry in systems, something that is also true in the current case: The model as presented and analyzed is formulated in an entirely symmetrical way, which means that a Pitchfork bifurcation is de facto built into it.

As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system. It is absolutely necessary to investigate the robustness of the results under scenarios where the symmetry is broken. The eco-evolutionary aspects of the model have currently no role in the observed dynamics, because they hinge on the symmetrical formulation.

I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations. Another potentially highly interesting element that has not been studied here is the potential for alternative stable states based on the density-dependent straying rate (subsection b in the Model formulation). The study currently addresses a scenario where the potential for population collapse due to collective navigation is ignored. I see that as the interesting angle; how do the dynamics caused by density-dependent straying interact with the dynamics caused by selective forces as imposed by the different environments on recruitment?

This manuscript can be improved by correcting the bifurcation analysis and by extending beyond perfect symmetry. The authors take a reasonable approach, using numerical analysis, which is fine, and I like the idea to focus on population robustness.  I understand that recruitment and population growth are not explicitly linked to environmental resources for reasons of mathematical simplicity. At the same time, I would prefer to see resources as an explicit variable in this model. This suggestion could directly be used to introduce asymmetries between two different locations and may be a feasible way to address the issues I raised above.

I would be looking forward to see a revised version of this manuscript that addresses my points.

It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?). As it is, I have no idea about the numerical techniques that were used or even whether the authors have studied both increasing and decreasing parameter values, as necessary to enable detection of alternative stable states and discontinuities.

Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.

I could not understand section c, Habitat heterogeneity in the Model formulation. In particular the paragraph on lines 139-144 should be explained and justified better, where the authors should make sure to explain the ‘integrate the two variables’ part more formally and explicitly.

As I write above, I appreciate the focus on robustness and through numerical analysis. At the same time this model would lend itself to some analysis too, even making some simplifying assumptions or investigating some boundary conditions. That would strengthen the insights from the results and the conclusions based on them. This would also possibly facilitate very straightforward validation of the kind of bifurcation point that is encountered (although the figures clearly show it cannot be a Fold bifurcation).

Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters. Right now I never found an explicit definition of NT, for example.

\subsection*{Minor comments}
I encountered some sloppiness in various parts of the manuscript, for example figure 3 is missing axes labels and has inconsistencies with the main text and caption. I did not find the SI figures or text, only the figure captions. Figure 4 was referenced before figure 3, as was the case with figures S2 and S1. Please see below for a list of textual errors that should be addressed (preferable before review, through using a spell checker). Note that these are the ones I could quickly jot down, but there are more.
Typos/errors on lines:
\begin{itemize}
\item 67:   ‘This model’ does not link back to anything about a model
\item 72:   determining
\item 90:   ‘than’ used awkwardly
\item 92:   inconsistency in number between noun and verb (equation that determine…)
\item 103  ‘is’ missing
\item 133: thgat
\item 139: differnce
\item 153:  inconsistency in number between noun and verb
\end{itemize}

\vspace{5mm}

\singlespacing
\closing{Sincerely,\\
\fromsig{\includegraphics[scale=0.2]{signature.jpg}}\\
\fromname{
Justin D. Yeakel\\
Assistant Professor\\
University of California, Merced}
}

\end{letter}
\end{document}

### 2017-09-06 Manuscript/FinalDraft_rev/msv6.tex
?3

\usepackage[font={small}]{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{xr}
\externaldocument[extfig-]{supp}
%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
% \usepackage{lineno}

@@

\bibliographystyle{prsb}

@@

\begin{abstract} %250 words
The spatial dispersal of individuals is known to play an important role in the dynamics of populations, and is central to metapopulation theory.
At the same time, local adaptation to different environmental conditions creates a geographic mosaic of evolutionary forces, where the combined drivers of selection and gene flow interact.
Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, potentially lowering the mean fitness of the recipient population.
Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, and recently proposed to be shaped by social interactions consistent with collective movement. %possibly emerging as a consequence of collective migration. 
The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
We show that intermediate levels of straying result in large gains in robustness, and that increased differences between sites (increased habitat heterogeneity) promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
\end{abstract} %250 words

@@

@@

\begin{equation}
\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
\label{eq:pe}
\end{equation}

\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
Portfolio effects greater than unity corresponds to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.

%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate biomass $N_T$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.

@@

Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
The fold bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus additional analyses on that range.

%Critical slowing down
As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).

@@

Straying had non-linear impacts on the recovery time of populations. 
When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{extfig-fig:relaxtraj_hdlh}).
Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{extfig-fig:relaxtraj_ldlh}).
When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\

@@

%Recovery times
Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
%Recovery times
%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{extfig-fig:relax_highh}).
% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.

%Lowering Phenotypic diversity with increasing straying rate
Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
Hence straying counters the effect of diversifying local adaptation. 
% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened. % with decreased $h^2$.
Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population. 
\\

\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{extfig-fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven. %I've checked this - it's a theshold type (logistic) response Supp fig?
The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness. %regardless of heritability.
%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{extfig-fig:relaxtheta}).
For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{extfig-fig:relaxtheta}).
These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).

If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{extfig-fig:inertia}).
This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{extfig-fig:mthetamvm}).

@@

The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady state and transient dynamics.
We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence. 
We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.

>>>>

\usepackage[font={small}]{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}
% \usepackage{xr}
% \externaldocument[extfig-]{supp}
%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
% \usepackage{lineno}

@@

% \bibliographystyle{prsb}

@@

\begin{abstract} %250 words
The spatial dispersal of individuals is known to play an important role in the dynamics of populations, and is central to metapopulation theory.
At the same time, local adaptation to environmental conditions creates a geographic mosaic of evolutionary forces, where the combined drivers of selection and gene flow interact.
Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats host to different environmental conditions, potentially lowering the fitness of the recipient population.
Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, and recently proposed to be shaped by social interactions consistent with collective movement. %possibly emerging as a consequence of collective migration. 
The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
We show that intermediate levels of straying result in large gains in robustness, and that increased habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
\end{abstract}

@@

\centerline{Salmon metapopulations, Straying, Dispersal, Eco-evolutionary dynamics, Alternative stable states}
\vspace{2mm}
\noindent {\bf Media Summary} The rate at which individuals move within a metapopulation has large impacts on extinction risk. Among migratory species such as salmon, the rate of dispersal, or straying, may be density dependent due to collective decision-making. We explore an eco-evolutionary model of two populations connected by dispersal, where the productivity of each is defined by a trait complex subject to local selection. We show that intermediate straying rates result in large gains in robustness, and that density-dependent straying promotes robustness. These results have important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.\\

@@

\begin{equation}
\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i^*)}}{{\rm E}(N_i^*)}\cdot \frac{{\rm E}(N_T^*)}{\sqrt{{\rm VAR}(N_T^*)}},
\label{eq:pe}
\end{equation}

\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
The steady state condition is denoted by `$*$'.
As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
Portfolio effects greater than unity corresponds to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.

%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady state biomass $N_T^*$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.

@@

Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N^*$.
The fold bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus additional analyses on that range.

%Critical slowing down
As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).

@@

Straying had non-linear impacts on the recovery time of populations. 
When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\

@@

%Recovery times
Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
%Recovery times
%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.

%Lowering Phenotypic diversity with increasing straying rate
Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
Hence straying counters the effect of diversifying local adaptation. 
% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened. % with decreased $h^2$.
Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population. 
\\

\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven. %I've checked this - it's a theshold type (logistic) response Supp fig?
The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness. %regardless of heritability.
%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).

If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).

@@

The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady state and transient dynamics.
We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence. 
We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
\\ \\
